A Tale of Two Dictionary Learning Problems

### 2020-08-13 manuscripts/thesis/discussion.tex
redid abstract

Another interesting remaining challenge is to work out for which special cases it is efficient to check that a solution to Prob.~\ref{InverseProblem} or \ref{OptimizationProblem} is unique up to noise and inherent ambiguities. Considering that the sufficient conditions detailed here are in general NP-hard to compute, are the necessary conditions also hard to compute? Are Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} then also hard (e.g., see \cite{tillmann2015computational})? Since Prob.~\ref{SLCopt} is intractable in general (i.e. including the noiseless case), but efficiently solvable by convex relaxation when the matrix $\mathbf{A}$ is known and has a large enough lower bound over sparse domains \cite{eldar2012compressed}, is there a version of Thm.~\ref{SLCopt} that lays down general conditions under which Prob.~\ref{OptimizationProblem} can be solved efficiently in full by similar means?  % [*** Incorporate this open problem with more words ***]

@@

I briefly expand on some of these directions below. It is my hope that these remaining challenges pique the interest of the community, and that practical guidelines can be established using the theoretical tools showcased here to support researchers applying sparse coding techniques in their data analysis pipelines.

@@

A commonly applied workaround to the intractability (see \cite{tillmann2015computational}) of Prob.~\ref{SLCopt} is to replace the norm $\|\mathbf{x}\|_0$ in \eqref{minsum} with its best convex approximation, $\|\mathbf{x}\|_1$, thereby transforming the inference of sparsest $\mathbf{\overline x}_i$ for fixed $\mathbf{B}$ into a convex optimization problem solvable by a linear program. Recalling the linear model \eqref{LinearModel} of data $\mathbf{z}_i$, the approach is to solve the following problem instead of attempting to solve Prob~\ref{SLCopt} directly.:

@@

A major advance in compressive sensing was the discovery that, when the matrix $\mathbf{B}$ is held fixed, the vectors $\mathbf{\overline x}_i$ that solve \eqref{l1minsum} also in fact minimize \eqref{minsum} provided $L_{2k}(\mathbf{B})$ is large enough \cite{eldar2012compressed}. Interestingly, the current work provides conditions on the generating dictionary $\mathbf{A}$ and $k$-sparse codes $\mathbf{x}_i$ under which \emph{all} matrices $\mathbf{B}$ solving Prob.~\ref{OptimizationProblem} have $L_{2k}(\mathbf{B})$ bounded from below; specifically, $L_{2k}(\mathbf{B}) \geq \left(L_{2k}(\mathbf{A}) - 2\eta C_1 \right) / \|\mathbf{D}\|_1$ in the case where $\mathbf{A}$ satisfies the spark condition \eqref{SparkCondition}. Thus, for suitable $\mathbf{A}$ there is some noise bound inside of which all solutions to Prob.~\ref{OptimizationProblem} are solutions to Prob.~\ref{ConvexifiedOptimizationProblem} as well. It is an open question as to whether there exist practical constraints which reject solutions to Prob.~\ref{ConvexifiedOptimizationProblem} that don't also solve Prob.~\ref{OptimizationProblem}.

@@

\begin{proposition}
If $\mathbf{A}$ has $L_2(\mathbf{A}) > 0$ and unit length-columns, and $\mathbf{x}_i = c_i\mathbf{e}_i$ with $c_i \neq 0$ for $i = 1, \ldots, m$, then every solution to Prob.~\ref{ConvexifiedOptimizationProblem} with matrices $\mathbf{B}$ having $m$ unit-length columns satisfies $\mathbf{A} = \mathbf{BP}$ and $\mathbf{x}_i = \mathbf{P}^\top \mathbf{\overline x}_i$ for some $m \times m$ permutation matrix $\mathbf{P}$.
\end{proposition}

\begin{proof}
We will show that the $\mathbf{\overline x}_i$ must in fact be $1$-sparse and then apply Thm.~\ref{DeterministicUniquenessTheorem}. Fix $i \in [m]$. Writing $\mathbf{\overline x}_i = \sum_{j=1}^m \overline c_{j} \mathbf{e}_j$, we have:
\begin{align}\label{gtc}
c_i = \|c_i\mathbf{Ae}_i\|_2 = \|\mathbf{B \overline x}_i\|_2 = \|\sum_{j=1}^m \overline c_{j} \mathbf{Be}_j\|_2 \leq \sum_{j=1}^m |\overline c_{j}| \|\mathbf{Be}_j\|_2 = \|\mathbf{\overline x}_i\|_1
\end{align}
So $\|\mathbf{\overline x}_i\|_1 \geq c_i$ for all $i \in [m]$. Therefore $\sum_{i=1}^m \|\mathbf{\overline x}_i\|_1 \geq \sum_i c_i$. But since $\mathbf{A}$ and the $\mathbf{x}_i$ satisfy the constraints of the minimization problem, we must also have $\sum_{i=1}^m \|\mathbf{\overline x}_i\|_1 \leq \sum_{i=1}^m \|\mathbf{x}_i\|_1 = \sum_i c_i$. Thus $\sum_{i=1}^m \|\mathbf{\overline x}_i\|_1 = \sum_i c_i$. Since again $\|\mathbf{\overline x}_i\|_1 \geq c_i$ for all $i \in [m]$, we must have $\| \mathbf{\overline x}_i\|_1 = c_i$ for all $i \in [m]$.

Revisiting \eqref{gtc}, we therefore have $c_i = \|\mathbf{B\overline x}_i\|_2 \leq \|\mathbf{\overline x}_i\|_1 = c_i$ for all $i \in [m]$, with equality holding only when $\overline c^{(i)}_j \mathbf{Be}_j$ are colinear. This would be the case either if the $\mathbf{\overline x}_i$ are $1$-sparse, in which case we may apply Thm.~\ref{DeterministicUniquenessTheorem} to guarantee both dictionary and code recovery, or if $\mathbf{B}$ has colinear columns. We can rule out the latter case because if that were so, we could consolidate all sets of colinear columns to form a dictionary with fewer than $m$ linearly independent columns and yet for which there must exist $1$-sparse $\mathbf{\overline x}_i$ solving Prob.~\ref{ConvexifiedOptimizationProblem}; but this would contradict the fact that $L_2(\mathbf{A}) > 0$. 
\end{proof}

%\begin{conjecture}\label{l1conjecture}
%Fix a matrix $\mathbf{A}$ with $\|\mathbf{Ae}_i\|_2 = 1$ for $i \in [m]$ and vectors $\mathbf{x}_i$ satisfying the assumptions of Thm.~\ref{SLCopt}. Every solution to Prob.~\ref{ConvexifiedOptimizationProblem} (with $\eta = \varepsilon/2$) and $\|\mathbf{Be}_i\|_2 = 1$. satisfies recovery guarantees \eqref{Cstable} and \eqref{b-PDa} when the corresponding bounds on $\eta$ are met.
%\end{theorem}

>>>>

Another interesting remaining challenge is to work out for which special cases it is efficient to check that a solution to Prob.~\ref{InverseProblem} or \ref{OptimizationProblem} is unique up to noise and inherent ambiguities. Considering that the sufficient conditions detailed here are in general NP-hard to compute, are the necessary conditions also hard to compute? Are Probs.~\ref{InverseProblem} and \ref{OptimizationProblem} then also hard (e.g., see \cite{tillmann2015computational})? Since Prob.~\ref{SLCopt} is intractable in general (i.e. including the noiseless case), but solvable in practice by convex relaxation when the matrix $\mathbf{A}$ is known and has a large enough lower bound over sparse domains \cite{eldar2012compressed}, is there a version of Thm.~\ref{SLCopt} that lays down general conditions under which Prob.~\ref{OptimizationProblem} can be solved efficiently in full by similar means?  % [*** Incorporate this open problem with more words ***]

@@

I briefly expand on some of these directions below. It is my hope that these remaining challenges pique the interest of the community and that the theoretical tools showcased here can be applied to derive practical dictionary learning guidelines for sparse representation modeling of real data.

@@

A commonly applied workaround to the intractability (see \cite{tillmann2015computational}) of Prob.~\ref{SLCopt} is to replace the $\ell_0$-norm in \eqref{minsum} with its best convex approximation, the $\ell_1$-norm, thereby transforming the inference of sparsest $\mathbf{\overline x}_i$ for fixed $\mathbf{B}$ into a convex optimization problem solvable by a linear program. Recalling the linear model \eqref{LinearModel} of data $\mathbf{z}_i$, the approach is to solve the following problem rather than solving Prob~\ref{SLCopt} directly.:

@@

A major advance in compressive sensing was the discovery of practical criteria guaranteeing that, when the matrix $\mathbf{B}$ is held fixed, the vectors $\mathbf{\overline x}_i$ that solve \eqref{l1minsum} also in fact solve \eqref{minsum} provided $L_{2k}(\mathbf{B})$ is large enough \cite{eldar2012compressed}. Interestingly, the current work provides conditions on the generating dictionary $\mathbf{A}$ and $k$-sparse codes $\mathbf{x}_i$ under which \emph{all} matrices $\mathbf{B}$ solving Prob.~\ref{OptimizationProblem} have $L_{2k}(\mathbf{B})$ bounded from below; specifically, $L_{2k}(\mathbf{B}) \geq \left(L_{2k}(\mathbf{A}) - 2\eta C_1 \right) / \|\mathbf{D}\|_1$ in the case where $\mathbf{A}$ satisfies the spark condition \eqref{SparkCondition}. Thus, for suitable $\mathbf{A}$ there should be some noise bound inside of which all solutions to Prob.~\ref{OptimizationProblem} are solutions to Prob.~\ref{ConvexifiedOptimizationProblem} as well. It is an open question as to whether there exist practical constraints which reject solutions to Prob.~\ref{ConvexifiedOptimizationProblem} that don't also solve Prob.~\ref{OptimizationProblem}.

@@

\begin{proposition}
If the matrix $\mathbf{A}$ has unit length-columns with $L_2(\mathbf{A}) > 0$ and $\mathbf{x}_i = c_i\mathbf{e}_i$ with $c_i \neq 0$ for $i = 1, \ldots, m$, then every solution to Prob.~\ref{ConvexifiedOptimizationProblem} with $\mathbf{B}$ also having $m$ unit-length columns satisfies $\mathbf{A} = \mathbf{BP}$ and $\mathbf{x}_i = \mathbf{P}^\top \mathbf{\overline x}_i$ for some $m \times m$ permutation matrix $\mathbf{P}$.
\end{proposition}

\begin{proof}
We will show that the $\mathbf{\overline x}_i$ must in fact be $1$-sparse and then apply Thm.~\ref{DeterministicUniquenessTheorem}. Fixing $i \in [m]$ and writing $\mathbf{\overline x}_i = \sum_{j=1}^m \overline c_{j} \mathbf{e}_j$, we have:
\begin{align}\label{gtc}
c_i = \|\mathbf{Ax}_i\|_2 = \|\mathbf{B \overline x}_i\|_2 = \|\sum_{j=1}^m \overline c_{j} \mathbf{Be}_j\|_2 \leq \sum_{j=1}^m |\overline c_{j}| \|\mathbf{Be}_j\|_2 = \|\mathbf{\overline x}_i\|_1
\end{align}
So $\|\mathbf{\overline x}_i\|_1 \geq c_i$ for all $i \in [m]$, and we have $\sum_{i=1}^m \|\mathbf{\overline x}_i\|_1 \geq \sum_{i=1}^m  c_i$. But since $\mathbf{A}$ and the $\mathbf{x}_i$ satisfy the constraints of the minimization problem, we must also have $\sum_{i=1}^m \|\mathbf{\overline x}_i\|_1 \leq \sum_{i=1}^m \|\mathbf{x}_i\|_1 = \sum_{i=1}^m  c_i$. Thus, $\sum_{i=1}^m \|\mathbf{\overline x}_i\|_1 = \sum_{i=1}^m  c_i$. Since again $\|\mathbf{\overline x}_i\|_1 \geq c_i$, it must be the case that in fact $\| \mathbf{\overline x}_i\|_1 = c_i$ for all $i \in [m]$.

Revisiting \eqref{gtc}, for a given $i \in [m]$ we have $c_i = \|\mathbf{B\overline x}_i\|_2 \leq \|\mathbf{\overline x}_i\|_1 = c_i$, with equality holding only when the vectors $\overline c_j \mathbf{Be}_j$ are colinear. This would be the case either if $\mathbf{\overline x}_i$ is $1$-sparse, or if $\mathbf{B}$ has colinear columns. We can rule out the latter case because if that were so, we could consolidate all sets of colinear columns to form a dictionary with fewer than $m$ columns and yet for which there must exist $1$-sparse $\mathbf{\overline x}_i$, $i = 1, \ldots, m$ solving Prob.~\ref{ConvexifiedOptimizationProblem}; but this would contradict the fact that $L_2(\mathbf{A}) > 0$. So it must be the case that in fact the $\mathbf{\overline x}_i$ are all $1$-sparse, and we may apply Thm.~\ref{DeterministicUniquenessTheorem}.
\end{proof}

The approach here was to show that the points $\mathbf{B \overline x}_i$ lie on a polytype which may identify with all of the data $\mathbf{z}_i = c_i\mathbf{Ae}_i$ only at its vertices, and that the $\mathbf{\overline x}_i$ must therefore be $1$-sparse. It seems reasonable to believe similar arguments would apply in the general case $k < m$, where the points $\mathbf{B \overline x}_i$ may align with all of the data only at the $k-1$-dimensional boundaries of the polytope. 
%As a next step toward this general case, I propose the following simple conjecture:

%\begin{conjecture}\label{l1conjecture}
%Fix a matrix $\mathbf{A}$ with unit-length columns and vectors $\mathbf{x}_i \in \mathbb{R}^m$ with $\|\mathbf{x}_i\|_1 = c$ $(c > 0)$ for all $i \in [m]$ satisfying the assumptions of Thm.~\ref{SLCopt}. Assume data $\mathbf{z}_i = \mathbf{Ax}_i$ for all $i \in [m]$. Every solution to Prob.~\ref{ConvexifiedOptimizationProblem} (with $\eta = 0$) for which $\mathbf{B}$ also has $m$ unit-length columns satisfies recovery guarantees \eqref{Cstable} and \eqref{b-PDa}.
%\end{conjecture}

### 2020-08-13 manuscripts/thesis/results.tex
redid abstract

\begin{theorem}\label{SLCopt}
Fix a matrix $\mathbf{A}$ and vectors $\mathbf{x}_i$ satisfying the assumptions of Thm.~\ref{DeterministicUniquenessTheorem}, only now with over \mbox{$(k-1)\left[ {\overline m \choose k} + |\mathcal{H}|k{\overline m \choose k-1}\right]$} vectors supported in general linear position in each $S \in \mathcal{H}$. Every solution to Prob.~\ref{OptimizationProblem} (with $\eta = \varepsilon/2$) satisfies recovery guarantees \eqref{Cstable} and \eqref{b-PDa} when the corresponding bounds on $\eta$ are met.
\end{theorem}

>>>>

\begin{theorem}\label{SLCopt}
Fix a matrix $\mathbf{A}$ and vectors $\mathbf{x}_i$ satisfying the assumptions of Thm.~\ref{DeterministicUniquenessTheorem}, only now with over \mbox{$(k-1)\left[ {\overline m \choose k} + |\mathcal{H}|k{\overline m \choose k-1}\right]$} vectors supported in general linear position in each $S \in \mathcal{H}$. Every solution to Prob.~\ref{OptimizationProblem} (with $\eta = \varepsilon/2$) satisfies the dictionary recovery and code recovery guarantees of Thm.~\ref{DeterministicUniquenessTheorem} when the corresponding bounds on $\varepsilon$ are met.
\end{theorem}
